Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-14 main.tex
proof

\newcommand{\sysS}{\mathsf{S}}
\newcommand{\Deri}{\Phi}

@@

@@

\subsection{Skew Bifibrations}
\lutz{$\phi$ instead of $f$}
A graph homomorphism $f: \tuple{\vG, \eG} \rightarrow \tuple{\vH, \eH}$ is a \bfit{fibration} if
for all $v \in \vG$ and $wf(v) \in \eH$, there exists a unique $\tilde{w} \in
\vG$ such that $\tilde{w}v \in \eG$ and $f(\tilde{w}) = w$, and is a \bfit{skew fibration}
if for all $v \in \vG$ and $wf(v) \in \eH$ there exists $\tilde{w} \in \vG$ such that
$\tilde{w}v \in \eG$ and $f(\tilde{w})w \notin \eH$.
A directed graph homomorphism is a \bfit{fibration} if \todo{}

@@

\begin{definition}
  Let $\gG$ and $\gH$ be forgraphs. A \bfit{skew bifibration} $\phi\colon\gG\to\gH$ is a fograph homomorphism that is a skew fibration on $\tuple{\vG, \eG} \to \tuple{\vH, \eH}$ and a fibration on the binding graphs $\bG\to\bH$.
  %etween fographs is a label-preserving and
%existential-preserving fograph homomorphism whose restriction to the
%binding graphs is a fibration.
\end{definition}

@@

\begin{lemma}\label{lem:shallow}
  Let $\vlinf{\rr}{}{\Scons B}{\Scons A}$ be an inference rule in
  $\FOMLS$. Then the sequent $\sqn{\cneg A,B}$ is provable in $\FOMLL$.
\end{lemma}
\juihsuan{For this, we need the introduction rule of $\fff$ in $\FOMLL$
(consider the $\aiD$ rule). But if we add it, we'll need the
$\mathsf{ai}\downarrow$ rule to have Lemma \ref{lem:MLL1->MLS1}}
\lutz{shit. so, we need another way to prove lemma~\ref{lem:MLS1->MLL1}, I'd like to avoid $\fff$-intro.}
\juihsuan{This lemma holds for all the rules except $\aiD$ (we may need to
mention that $A \fequ B \Rightarrow \ \vdash_{\FOMLL} \dual{B}, A$). The inductive
step for $\aiD$ in the proof of Lemma \ref{lem:MLS1->MLL1} can be solved by
replacing $t$ with $a \vlor \dual{a}$ and the $\ttt$-rule with the $\ax$
 rule followed by the $\vlor$ rule}\lutz{yes! and similar for $\forall$. Can you do that?}
\begin{proof}
  This is a straightforward exercise that we leave to the reader.
\end{proof}

@@

\begin{proof}
  We proceed by induction on the number of inference steps in the
  proof of $C$ in $\FOMLS$. Consider the bottommost rule instance
  $\vlinf{\rr}{}{\Scons B}{\Scons A}$. By induction hypothesis we have
  a $\FOMLL$ proof of $\sqn{\Scons A}$, and by
  Lemma~\ref{lem:context}, we have a $\FOMLL$ proof of
  $\sqn{\cneg{\Scons A},\Scons B}$. We can compose them via $\cutr$:
  \begin{equation*}
    \vliinf{\cutr}{}{\sqn{\Scons B}}{\sqn{\Scons A}}{\sqn{\cneg{\Scons
          A},\Scons B}}
  \end{equation*}
  and then apply Theorem~\ref{thm:cutelim}.
\end{proof}

@@

In this paragraph, we associate to each formula $A$ its \bfit{formula tree}
$\formtree{A}$, a directed
tree with leaves labelled by atoms, internal nodes labelled by connectives and
quantifiers, and edges directed from leaves to the root.

Let $\Gamma$ be a sequent in $\FOMLL$. Consider the formula tree
$\formtree{\form{\Gamma}}$. A \bfit{link} on $\Gamma$ is a pair of
leaves whose atoms are pre-dual. A \bfit{linking} $\lambda$ on $\Gamma$ is a set
of
disjoint links such that each leaf of $\Gamma$ is either labelled by $\ttt$ or
in exactly one link. Similar to the set of links in linked fographs, a linking can be seen as a
unification problem, and a dualizer of the linking $\lambda$ is an assignment
unifying all the links in $\lambda$. There exists a \bfit{most general dualizer}
 of $\lambda$ if $\lambda$ has a dualizer. \juihsuan{Now I use the same terminology as for linked fographs}\lutz{use $\delta$ for the dualizer (or even better, make it a macro)}
A \bfit{dependency} is a pair $(\single \exists x, \single \forall y)$ of nodes
such that the most general dualizer assigns to $x$ a term containing $y$.

Let $\lambda$ is a linking on $\Gamma$ that has a dualizer.
The \bfit{unification structure} $\unifstr{\lambda}$
associated to $\lambda$ is the formula tree $\formtree{\form{\Gamma}}$ together with an undirected
edge between leaves $l$ and $l'$ for every link $\{l, l'\}$ in $\lambda$ and a
directed edge from $\single \exists x$ to $\single \forall y$ for every
dependency $(\single \exists x, \single \forall y)$.

@@

\begin{thm}
  Let $\gC$ be a fonet with linking $\linkingof\gC$ and dualizer
  $\delta$, and let $A$ be a formula with $\graphof A=\gC$.  Then
  there is a derivation $\vlder{\FOMLS}{\Deri}{A}{\ttt}$ such that
  $\linkingof\gC=\linkingof{\Deri}$ and $\delta=\substof{\Deri}$.
\end{thm}

>>>>

\newcommand{\sysS}{\mathsf{S}}
\newcommand{\Deri}{\Phi}
\newcommand{\Proof}{\Pi}

@@

\newcommand{\dualizer}{\delta}

@@

\subsection{Skew Bifibrations}
\lutz{$\phi$ instead of $f$}
A graph homomorphism $\phi\colon \tuple{\vG, \eG} \rightarrow \tuple{\vH, \eH}$ is a \bfit{fibration} if
for all $v \in \vG$ and $w\phi(v) \in \eH$, there exists a unique $\tilde{w} \in
\vG$ such that $\tilde{w}v \in \eG$ and $\phi(\tilde{w}) = w$, and is a \bfit{skew fibration}
if for all $v \in \vG$ and $w\phi(v) \in \eH$ there exists $\tilde{w} \in \vG$ such that
$\tilde{w}v \in \eG$ and $\phi(\tilde{w})w \notin \eH$.
A directed graph homomorphism is a \bfit{fibration} if \todo{}

@@

\begin{definition}
  Let $\gG$ and $\gH$ be fographs. A \bfit{skew bifibration} $\phi\colon\gG\to\gH$ is a fograph homomorphism that is a skew fibration on $\tuple{\vG, \eG} \to \tuple{\vH, \eH}$ and a fibration on the binding graphs $\bG\to\bH$.
  %etween fographs is a label-preserving and
%existential-preserving fograph homomorphism whose restriction to the
%binding graphs is a fibration.
\end{definition}

@@

\begin{lemma}\label{lem:shallow}
  Let $\vlinf{\rr}{}{\Scons B}{\Scons A}$ be an inference rule in
  $\FOMLS$ other than $\aiD$ and $\forall$. Then the sequent $\sqn{\cneg A,B}$ is provable in $\FOMLL$.
\end{lemma}
\juihsuan{For this, we need the introduction rule of $\fff$ in $\FOMLL$
(consider the $\aiD$ rule). But if we add it, we'll need the
$\mathsf{ai}\downarrow$ rule to have Lemma \ref{lem:MLL1->MLS1}}
\lutz{shit. so, we need another way to prove lemma~\ref{lem:MLS1->MLL1}, I'd like to avoid $\fff$-intro.}
\juihsuan{This lemma holds for all the rules except $\aiD$ (we may need to
mention that $A \fequ B \Rightarrow \ \vdash_{\FOMLL} \dual{B}, A$). The inductive
step for $\aiD$ in the proof of Lemma \ref{lem:MLS1->MLL1} can be solved by
replacing $t$ with $a \vlor \dual{a}$ and the $\ttt$-rule with the $\ax$
 rule followed by the $\vlor$ rule}\lutz{yes! and similar for $\forall$. Can you do that?}
\begin{proof}
  This is a straightforward exercise that we leave to the reader.
\end{proof}

@@

$\vlinf{\faD}{}{\Scons{\forall x.\ttt}}{\Scons\ttt}$

$\vlinf{\axr}{}{\sqn{a,\cneg a}}{}
            \qquad
            \vlinf{\ttt}{}{\sqn{\ttt}}{}
$

\begin{proof}
  We proceed by induction on the number of inference steps in the
  proof of $C$ in $\FOMLS$. Consider the bottommost rule instance
  $\vlinf{\rr}{}{\Scons B}{\Scons A}$. By induction hypothesis we have
  a $\FOMLL$ proof $\Proof$ of $\sqn{\Scons A}$.
  We now make a case analysis on $\rr$:
  \begin{itemize}
    \setlength\itemsep{.5em}
    \item $\vlinf{\aiD}{}{\Scons{a\vlor\cneg a}}{\Scons\ttt}$ :
      In this case, we replace simultaneously in $\Pi$ the corresponding
      occurrences of $\ttt$ with $a \vlor \cneg a$ and the rule instance
      $\vlinf{\ttt}{}{\sqn{\ttt}}{}$ with the derivation
      $
        \vlderivation{
          \vlin{\vlor}{}{\sqn{a \vlor \cneg a}}{
            \vlin{\axr}{}{\sqn{a, \cneg a}}{
              \vlhy{}}}}
      $, and we get a proof $\Pi'$ of $\sqn{\Scons{a \vlor \cneg a}}$.
    \item $\vlinf{\faD}{}{\Scons{\forall x.\ttt}}{\Scons\ttt}$
      In this case, we replace simultaneously in $\Pi$ the corresponding
      occurrences of $\ttt$ with $\forall x.\ttt$ and the rule instance
      $\vlinf{\ttt}{}{\sqn{\ttt}}{}$ with the derivation
      $
      \vlderivation{
        \vlin{\forall}{}{\sqn{\forall x.\ttt}}{
          \vlin{\axr}{}{\sqn{\ttt}}{
            \vlhy{}}}}
      $, and we get a proof $\Pi'$ of $\sqn{\Scons{\forall x.\ttt}}$.
    \item $\rr \notin \set{\aiD, \faD}$ :
      By Lemma~\ref{lem:context}, we have a $\FOMLL$ proof of
      $\sqn{\cneg{\Scons A},\Scons B}$. We can compose them via $\cutr$:
      \begin{equation*}
        \vliinf{\cutr}{}{\sqn{\Scons B}}{\sqn{\Scons A}}{\sqn{\cneg{\Scons
          A},\Scons B}}
      \end{equation*}
      and then apply Theorem~\ref{thm:cutelim}.
  \end{itemize}
  \tocheck{}
  \end{proof}

@@

In this paragraph, we associate each formula $A$ with its \bfit{formula tree}
$\formtree{A}$, a directed
tree with leaves labelled by atoms, internal nodes labelled by connectives and
quantifiers, and edges directed from leaves to the root.

Let $\Gamma$ be a sequent in $\FOMLL$. Consider the formula tree
$\formtree{\form{\Gamma}}$. A \bfit{link} on $\Gamma$ is a pair of
leaves whose atoms are pre-dual. A \bfit{linking} $\lambda$ on $\Gamma$ is a set
of
disjoint links such that each leaf of $\Gamma$ is either labelled by $\ttt$ or
in exactly one link. Similar to the set of links in linked fographs, a linking can be seen as a
unification problem, and a \bfit{dualizer} $\dualizer$ of the linking $\lambda$ is an assignment
unifying all the links in $\lambda$. There exists a \bfit{most general dualizer}
 of $\lambda$ if $\lambda$ has a dualizer. \juihsuan{Now I use the same terminology as for linked fographs}\lutz{use $\delta$ for the dualizer (or even better, make it a macro)}
A \bfit{dependency} is a pair $(\single \exists x, \single \forall y)$ of nodes
such that the most general dualizer assigns to $x$ a term containing $y$.

Let $\lambda$ is a linking on $\Gamma$ that has a dualizer.
The \bfit{unification structure} $\unifstr{\lambda}$
associated with $\lambda$ is the formula tree $\formtree{\form{\Gamma}}$ together with an undirected
edge between leaves $l$ and $l'$ for every link $\{l, l'\}$ in $\lambda$ and a
directed edge from $\single \exists x$ to $\single \forall y$ for every
dependency $(\single \exists x, \single \forall y)$.

@@

\begin{thm}
  Let $\gC$ be a fonet with linking $\linkingof\gC$ and dualizer
  $\dualizer$, and let $A$ be a formula with $\graphof A=\gC$.  Then
  there is a derivation $\vlder{\FOMLS}{\Deri}{A}{\ttt}$ such that
  $\linkingof\gC=\linkingof{\Deri}$ and $\dualizer=\substof{\Deri}$.
\end{thm}
